---
title: "Finding balance"
comments: yes
date: '2015-12-18'
modified: `r format(Sys.time(), '%Y-%m-%d')`
layout: post
excerpt: "Post-hoc causal inference by matching"
published: no
status: draft
tags:
- R
- propensity score matching
- Matching
- logistic regression
- statistics
categories: Rstats
---

```{r,include = FALSE}

#library(checkpoint)
#checkpoint("2015-05-15")

library(dplyr)
library(Matching)

opts_chunk$set(warning = FALSE, message = FALSE)

```

In the ideal world, when we want to test the efficacy of a treatment against a control, we apply the treatment randomly to our sample.
In doing so, it shouldn't matter to the outcome of the experiment whether sample $n$ was assigned to the control or to the treatment group.
In effect we are assuming that the potential outcome of the treatment is dependent solely on the application of the treatment, and not on any underlying differences in our treatment groups.

Simple enough, but what do we do when we were not able to select randomly.
In my ecological research this was a fairly common occurence, and there are various statistical methods than we can employ to control for this issue.
On some occasions, we might have even less control than this, and our experiment is based on the application of a treatment on a non-random sample over which we had entirely no control.
Or (more likely: and) if we are dealing with human subjects, it would not have been ethical to design a randomised control experiment of this sort in the first place.

This is where the suite of matching techniques can be useful.
By looking at a range of covariates which describe the individuals within our experiments, we can start to match the individuals within our treatment group with individuals within the control, and 

Following Sekhon, we can formalise some of these


```{r}
sessionInfo()
```
